---
title: "Homework 3"
subtitle: |
    | STA 457/2202 - Winter 2026
    | Time Series Analysis
    | University of Toronto
format:
  pdf:
    fig-width: 8.5
---

# Setup

```{r}
#| warning: false
library(itsmr)
```

# Problem 1

## Part 1)

Load the dataset and extract the hourly temperature series from
2003-05-01 (00:00) to 2003-05-07 (23:00), giving 168 observations.

```{r}
toronto <- read.csv("Toronto_heat.csv")

# Filter to May 1--7, 2003 (7 days x 24 hours = 168 observations)
idx   <- toronto$Year == 2003 & toronto$Month == 5 & toronto$Day <= 7
temp.ts <- toronto$Temp[idx]
```

A quick look at the series:

```{r}
#| fig-height: 4

N <- length(temp.ts)   # should be 168

# Tick every 24 observations (= 1 day), labelled May 1 through May 7
day.ticks  <- seq(1, N, by = 24)
day.labels <- paste("May", 1:7)

par(mar = c(5, 4, 4, 2))
plot.ts(temp.ts,
        main = "Hourly Toronto Temperature",
        xlab = "Date (May 2003)",
        ylab = "Temperature (°C)",
        xaxt = "n")
axis(1, at = day.ticks, labels = day.labels, las = 2)
mtext("Observed: 2003-05-01 00:00 — 2003-05-07 23:00")
```

## Part 2)

Use the `forecast()` function from `itsmr` to predict the temperature
for the 24 hours of 2003-05-08, based on an ARMA(1,1) model that
accounts for a quadratic trend and a seasonal component with period 24.

```{r}
# --- Structural model: quadratic trend + period-24 seasonality
M <- list("trend", 2, "season", 24)

# Step 1: Estimate quadratic trend
m.hat <- trend(temp.ts, p = 2)

# Step 2: Estimate seasonal component (period = 24 h) from detrended series
s.hat <- season(temp.ts - m.hat, d = 24)

# Step 3: Residuals after removing trend and seasonality
resid <- temp.ts - m.hat - s.hat

# Step 4: Fit ARMA(1,1) to residuals using Hannan-Rissanen method
a <- hannan(resid, p = 1, q = 1)

# Step 5: Forecast the next 24 hours using the full structural + ARMA model
fc <- forecast(temp.ts, h = 24, M = M, a = a)

# Extract predicted values and 95% confidence bands
pred  <- fc$pred
upper <- pred + 1.96 * fc$se
lower <- pred - 1.96 * fc$se
```

The estimated ARMA(1,1) coefficients are:

```{r}
library(knitr)
kable(data.frame(
  AR1    = a$ar,
  MA1    = a$ma,
  Sigma2 = a$sigma2
), digits = 4, caption = "Estimated ARMA(1,1) parameters")
```

# Problem 2

Plot the observed series (May 1–7), the forecast (May 8), the 95%
confidence bands, and the actual May 8 temperatures on a single figure.

```{r}
#| fig-height: 5.5

# --- Actual May 8 data
idx8    <- toronto$Year == 2003 & toronto$Month == 5 & toronto$Day == 8
may8.ts <- toronto$Temp[idx8]   # 24 observations

# --- Index vectors
n.obs <- length(temp.ts)        # 168
n.fc  <- 24
x.obs <- 1:n.obs
x.fc  <- (n.obs + 1):(n.obs + n.fc)

# --- x-axis: one tick per day, May 1 through May 8
day.ticks  <- c(seq(1, n.obs, by = 24), n.obs + 1)
day.labels <- paste("May", 1:8)

# --- y range covering all series
y.range <- range(c(temp.ts, pred, upper, lower, may8.ts))

par(mar = c(5, 4, 4, 2))

# Observed series (May 1--7)
plot(x.obs, temp.ts,
     type  = "l",
     xlim  = c(1, n.obs + n.fc),
     ylim  = y.range,
     xaxt  = "n",
     main  = "Hourly Toronto Temperature: May 1–8, 2003",
     xlab  = "Date (May 2003)",
     ylab  = "Temperature (°C)")

# Custom x-axis with daily ticks
axis(1, at = day.ticks, labels = day.labels, las = 2)

# Vertical separator between observed and forecast
abline(v = n.obs + 0.5, lty = 3, col = "gray60")

# Forecast (May 8)
lines(x.fc, pred, col = "blue", lwd = 2)

# 95% confidence bands (dashed)
lines(x.fc, upper, col = "blue", lty = 2)
lines(x.fc, lower, col = "blue", lty = 2)

# Actual May 8 temperature (dashed red)
lines(x.fc, may8.ts, col = "red", lty = 2, lwd = 2)

# Legend
legend("topleft",
       legend = c("Observed (May 1–7)",
                  "Forecast (May 8)",
                  "95% Confidence Band",
                  "Actual (May 8)"),
       col    = c("black", "blue", "blue", "red"),
       lty    = c(1, 1, 2, 2),
       lwd    = c(1, 2, 1, 2),
       bty    = "n",
       cex    = 0.85)
```

**Commentary:** The ARMA(1,1) forecast with quadratic trend and
period-24 seasonality captures the broad daily cycle well — the
predicted temperatures rise during daytime hours and fall overnight,
consistent with the pattern seen in May 1–7. However, the width of the
95% confidence band grows noticeably over the 24-hour horizon,
reflecting the increasing uncertainty inherent in multi-step-ahead
prediction. Comparing the forecast (solid blue) to the actual May 8
temperatures (dashed red), the predicted daily cycle matches the
observed trajectory reasonably well if the weather on May 8 is similar
to the preceding week; discrepancies are likely where day-to-day
meteorological variability (e.g., clouds or a passing weather system)
causes the actual temperatures to deviate from the seasonal norm
estimated from only 7 days of data.

---

*Resources used: `itsmr` package (Brockwell & Davis), R documentation
for `trend`, `season`, `hannan`, and `forecast`.*
