---
title: "Homework 3"
subtitle: |
    | STA457/2202 — Winter 2026
    | Time Series Analysis
format:
  pdf:
    fig-width: 8.5
---

## Setup

```{r setup}
#| warning: false
#| message: false
library(itsmr)
library(knitr)
```

---

# Problem 1

## Part 1: Create the Time Series (May 1–7, 2003)

```{r p1-ts}
toronto <- read.csv("Toronto_heat .csv")

# Extract hourly temperatures for May 1 (00:00) to May 7 (23:00)
may17  <- toronto[toronto$Month == 5 & toronto$Day %in% 1:7,  "Temp"]
n      <- length(may17)   # 168 observations (7 days × 24 hours)

head(may17)
```

## Part 2: Forecast Using ARMA(1,1) via `forecast()`

We decompose the series by removing a quadratic trend and a seasonal component
with period $d = 24$ hours, then fit ARMA(1,1) to the residuals and forecast
24 steps ahead using `forecast()` from `itsmr`.

```{r p1-forecast}
# --- Step 1: Remove quadratic trend (p = 2)  [itsmr: trend()]
m.trend <- trend(may17, p = 2)

# Also fit lm() to extrapolate the trend beyond the observed data
t.vec    <- 1:n
m.fit    <- lm(may17 ~ t.vec + I(t.vec^2))

# --- Step 2: Remove seasonal component, period = 24 hours  [itsmr: season()]
resid.m  <- may17 - m.trend
s.season <- season(resid.m, d = 24)

# --- Step 3: ARMA residuals
y.resid <- may17 - m.trend - s.season

# --- Step 4: Fit ARMA(1,1) to residuals, then specify model  [itsmr: specify()]
fit.arma <- arima(y.resid, order = c(1, 0, 1), include.mean = FALSE)
model    <- specify(ar     = fit.arma$coef["ar1"],
                    ma     = fit.arma$coef["ma1"],
                    sigma2 = fit.arma$sigma2)

kable(t(as.data.frame(model)), caption = "ARMA(1,1) Model Parameters")

# --- Step 5: Forecast 24 hours ahead  [itsmr: forecast()]
fc <- itsmr::forecast(y.resid, model, h = 24)

# --- Step 6: Reconstruct full temperature forecast
#   Project the quadratic trend forward to hours 169–192
t.future <- (n + 1):(n + 24)
m.future <- predict(m.fit, newdata = data.frame(t.vec = t.future))

#   Seasonal pattern repeats every 24 hours; hours 169–192 match hours 1–24
s.future <- s.season[1:24]

# Access forecast columns by index (robust across itsmr versions)
pred.temp  <- fc[, 1] + m.future + s.future
pred.lower <- fc[, 2] + m.future + s.future
pred.upper <- fc[, 3] + m.future + s.future
```

---

# Problem 2

```{r p2-plot}
#| fig-height: 5

# Actual May 8 temperatures (00:00 – 23:00)
may8   <- toronto[toronto$Month == 5 & toronto$Day == 8, "Temp"]
x.hist <- 1:n
x.pred <- (n + 1):(n + 24)

# y-axis range covering all series
ylim.range <- range(c(may17, pred.temp, pred.lower, pred.upper, may8))

# x-axis: one tick per day at midnight
day.ticks  <- seq(1, n + 24, by = 24)
day.labels <- paste0("May ", 1:9)

par(mar = c(5, 4, 4, 2))

# 1. Historical data: May 1–7
plot(x.hist, may17,
     type = "l",
     xlim = c(1, n + 24),
     ylim = ylim.range,
     main = "Toronto Hourly Temperature: May 1–8, 2003",
     xlab = "Date",
     ylab = "Temperature (°C)",
     xaxt = "n",
     col  = "black",
     lwd  = 1.5)

axis(1, at = day.ticks, labels = day.labels[seq_along(day.ticks)], las = 2)
abline(v = n + 0.5, lty = 3, col = "grey60")   # forecast start marker

# 2. Predicted values for May 8
lines(x.pred, pred.temp, col = "blue", lwd = 1.5)

# 3. Upper and lower 95% confidence bands (dashed, lty = 2)
lines(x.pred, pred.upper, col = "blue", lty = 2)
lines(x.pred, pred.lower, col = "blue", lty = 2)

# 4. Actual May 8 temperatures (dashed)
lines(x.pred, may8, col = "red", lty = 2, lwd = 1.5)

legend("topleft",
       legend = c("May 1–7 (observed)",
                  "May 8 (ARMA(1,1) forecast)",
                  "95% confidence band",
                  "May 8 (actual)"),
       col  = c("black", "blue", "blue", "red"),
       lty  = c(1, 1, 2, 2),
       lwd  = c(1.5, 1.5, 1, 1.5),
       cex  = 0.8)
```

**Commentary:** The ARMA(1,1) forecast captures the broad shape of May 8th's
temperature, including the overnight low and afternoon rise, thanks to the
quadratic trend and 24-hour seasonal component we removed. The actual values
mostly fall within the 95% confidence band for the first half of the day,
but the model underestimates the afternoon peak, which suggests some
day-to-day variability that cannot be fully captured by a simple ARMA(1,1)
residual model.

---

# Problem 3

Let $A_1, A_2$ be uncorrelated, zero-mean RVs with unit variance, and let
$\omega \in [0,\pi]$ be a fixed frequency. Define
$$
X_t = A_1\cos(\omega t) + A_2\sin(\omega t), \qquad t = 0, \pm 1, \ldots
$$

**Find $P_2 X_3$ and its mean squared error.**

**Step 1 — Autocovariance function.**

$$
\gamma(h) = \mathrm{Cov}(X_t, X_{t+h})
= \cos(\omega h)
$$

because $\mathrm{Var}(A_1) = \mathrm{Var}(A_2) = 1$ and $A_1 \perp A_2$.
In particular $\gamma(0) = 1$.

**Step 2 — Normal equations for $P_2 X_3 = a_1 X_1 + a_2 X_2$.**

$$
\begin{pmatrix} \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0) \end{pmatrix}
\begin{pmatrix} a_1 \\ a_2 \end{pmatrix}
=
\begin{pmatrix} \mathrm{Cov}(X_1, X_3) \\ \mathrm{Cov}(X_2, X_3) \end{pmatrix}
=
\begin{pmatrix} \gamma(2) \\ \gamma(1) \end{pmatrix}
$$

$$
\begin{pmatrix} 1 & \cos\omega \\ \cos\omega & 1 \end{pmatrix}
\begin{pmatrix} a_1 \\ a_2 \end{pmatrix}
=
\begin{pmatrix} \cos(2\omega) \\ \cos(\omega) \end{pmatrix}
$$

**Step 3 — Solve.** Let $c = \cos\omega$.

From the second equation: $a_2 = c - c\,a_1$.
Substituting into the first:

$$
a_1 + c(c - c\,a_1) = 2c^2 - 1
\;\Longrightarrow\;
a_1(1 - c^2) = -(1 - c^2)
\;\Longrightarrow\;
a_1 = -1
$$

Then $a_2 = c + c = 2\cos\omega$.

$$
\boxed{P_2 X_3 = 2\cos(\omega)\,X_2 - X_1}
$$

**Step 4 — MSE.**

$$
\mathrm{MSE} = \gamma(0) - \bigl(a_1\,\gamma(2) + a_2\,\gamma(1)\bigr)
= 1 - \bigl(-\cos(2\omega) + 2\cos^2\omega\bigr)
= 1 - 1 = \boxed{0}
$$

The MSE is **zero**: $X_t$ satisfies the exact recurrence
$X_t = 2\cos(\omega)\,X_{t-1} - X_{t-2}$, so it can be predicted perfectly
from two past values.

---

# Problem 4

Let $X_1, X_2, X_4, X_5$ be taken from the MA(1) model
$$
X_t = Z_t + \theta Z_{t-1}, \qquad \{Z_t\} \sim wn(0,\sigma^2).
$$

### Part a)

**Find $P(X_3 \mid X_4, X_5)$: the best linear predictor of $X_3$ in terms
of $X_4$ and $X_5$.**

The autocovariance function of MA(1) is
$$
\gamma(0) = \sigma^2(1+\theta^2),\quad
\gamma(1) = \sigma^2\theta,\quad
\gamma(h) = 0 \text{ for } |h| \geq 2.
$$

Write $P(X_3 \mid X_4, X_5) = a_4 X_4 + a_5 X_5$.
The normal equations are:
$$
\begin{pmatrix}\gamma(0) & \gamma(1)\\\gamma(1) & \gamma(0)\end{pmatrix}
\begin{pmatrix}a_4\\a_5\end{pmatrix}
=
\begin{pmatrix}\mathrm{Cov}(X_4,X_3)\\\mathrm{Cov}(X_5,X_3)\end{pmatrix}
=
\begin{pmatrix}\gamma(1)\\\gamma(2)\end{pmatrix}
=
\begin{pmatrix}\sigma^2\theta\\0\end{pmatrix}.
$$

From the second equation:
$$
a_4 = -\frac{(1+\theta^2)}{\theta}\,a_5.
$$

Substituting into the first and simplifying:
$$
a_5\cdot\frac{-(1+\theta^2)^2 + \theta^2}{\theta} = \theta
\;\Longrightarrow\;
a_5 = \frac{-\theta^2}{1+\theta^2+\theta^4}.
$$

Then:
$$
a_4 = \frac{\theta(1+\theta^2)}{1+\theta^2+\theta^4}.
$$

$$
\boxed{
P(X_3 \mid X_4, X_5)
= \frac{\theta(1+\theta^2)}{1+\theta^2+\theta^4}\,X_4
  - \frac{\theta^2}{1+\theta^2+\theta^4}\,X_5
}
$$

### Part b)

**Find the MSE of the predictor from Part a).**

$$
\mathrm{MSE}
= \gamma(0) - \bigl(a_4\,\gamma(1) + a_5\cdot 0\bigr)
= \sigma^2(1+\theta^2) - \frac{\theta(1+\theta^2)}{1+\theta^2+\theta^4}\cdot\sigma^2\theta
$$

$$
= \sigma^2(1+\theta^2)\left(1 - \frac{\theta^2}{1+\theta^2+\theta^4}\right)
= \sigma^2(1+\theta^2)\cdot\frac{1+\theta^4}{1+\theta^2+\theta^4}
$$

$$
\boxed{
\mathrm{MSE}
= \frac{\sigma^2(1+\theta^2)(1+\theta^4)}{1+\theta^2+\theta^4}
}
$$

---

# Problem 5 — Graduate Students Only

Let $\{Y_t\}$ be the ARMA(1,1) process:
$$
Y_t - \tfrac{1}{2}Y_{t-1} = W_t + \tfrac{1}{2}W_{t-1}, \qquad
W_t \sim wn(0,\sigma^2).
$$
So $\phi = \tfrac{1}{2}$, $\theta = \tfrac{1}{2}$.

**Autocovariance (textbook p. 78 formula):**

$$
\gamma(0)
= \frac{\sigma^2\bigl(1 + 2\phi\theta + \theta^2\bigr)}{1-\phi^2}
= \frac{\sigma^2\bigl(1 + \tfrac{1}{2} + \tfrac{1}{4}\bigr)}{\tfrac{3}{4}}
= \frac{\tfrac{7}{4}\sigma^2}{\tfrac{3}{4}}
= \frac{7\sigma^2}{3}
$$

$$
\gamma(1)
= \frac{(\phi+\theta)(1+\phi\theta)\,\sigma^2}{1-\phi^2}
= \frac{1\cdot\tfrac{5}{4}\,\sigma^2}{\tfrac{3}{4}}
= \frac{5\sigma^2}{3}
$$

### Part a)

**Find $P(Y_2 \mid Y_1)$.**

Since we are predicting one step ahead from a single observation:
$$
P(Y_2 \mid Y_1) = \frac{\gamma(1)}{\gamma(0)}\,Y_1
= \frac{5\sigma^2/3}{7\sigma^2/3}\,Y_1
= \frac{5}{7}\,Y_1
$$

$$
\boxed{P(Y_2 \mid Y_1) = \tfrac{5}{7}\,Y_1}
$$

### Part b)

**Compute the MSE of $P(Y_2 \mid Y_1)$.**

$$
\mathrm{MSE}
= \gamma(0) - \frac{[\gamma(1)]^2}{\gamma(0)}
= \frac{7\sigma^2}{3} - \frac{(5\sigma^2/3)^2}{7\sigma^2/3}
= \frac{7\sigma^2}{3} - \frac{25\sigma^2}{21}
= \frac{49\sigma^2 - 25\sigma^2}{21}
$$

$$
\boxed{\mathrm{MSE} = \frac{24\sigma^2}{21} = \frac{8\sigma^2}{7}}
$$
