---
title: "Homework 3"
subtitle: |
    | STA457/2202 - Winter 2026
    | Time Series Analysis | University of Toronto
    | Deadline: Friday, March 13th, 23:59
format:
  pdf:
    fig-width: 8.5
    fig-height: 5
---

# Setup

```{r setup}
#| warning: false
#| message: false
library(itsmr)
```

---

# Problem 1

## Part 1: Create the Time Series (May 1–7, 2003)

```{r p1-ts}
toronto <- read.csv("Toronto_heat .csv")

# Filter May 2003
may_data <- toronto[toronto$Year == 2003 & toronto$Month == 5, ]

# May 1–7: 7 days × 24 hours = 168 observations
may1to7 <- may_data[may_data$Day <= 7, "Temp"]
x <- ts(may1to7, frequency = 24)
n <- length(x)  # 168
```

## Part 2: Forecast Next 24 Hours via ARMA(1,1) with Trend + Seasonality

The model accounts for a **quadratic trend** and a **seasonal component** with period 24, with an ARMA(1,1) fitted to the residuals.

```{r p1-forecast}
# --- Step 1: Remove quadratic trend ---
t_idx <- 1:n
lm_trend <- lm(x ~ t_idx + I(t_idx^2))
m_fit <- fitted(lm_trend)
x_detrend <- x - m_fit

# --- Step 2: Remove seasonal component (period = 24 hours) ---
s_fit <- season(x_detrend, d = 24)
x_resid <- x_detrend - s_fit

# --- Step 3: Estimate ARMA(1,1) on residuals ---
M0 <- specify(ar = c(0.1), ma = c(0.1), sigma2 = var(x_resid))
M_est <- estimate(x_resid, M0)

# --- Step 4: Forecast 24 steps ahead ---
fc <- forecast(x_resid, M_est, h = 24)

# --- Step 5: Reconstruct predictions by adding back trend + seasonality ---
t_future <- (n + 1):(n + 24)

# Extrapolate quadratic trend
coefs <- coef(lm_trend)
m_future <- coefs[1] + coefs[2] * t_future + coefs[3] * t_future^2

# Seasonal pattern repeats with period 24
s_future <- s_fit[((t_future - 1) %% 24) + 1]

# Final forecast + confidence bands (±1.96 SE)
fc_pred  <- fc$pred + m_future + s_future
fc_upper <- fc$pred + 1.96 * fc$se + m_future + s_future
fc_lower <- fc$pred - 1.96 * fc$se + m_future + s_future
```

---

# Problem 2

```{r p2-plot}
# Actual May 8 data (for comparison)
may8 <- may_data[may_data$Day == 8, "Temp"]

y_range <- range(c(as.numeric(x), fc_upper, fc_lower, may8), na.rm = TRUE)

# Day-tick positions: one per day (every 24 obs)
day_ticks  <- seq(1, n + 24, by = 24)
day_labels <- paste("May", 1:8)

plot(1:n, as.numeric(x),
     type = "l", col = "black",
     xlim = c(1, n + 24), ylim = y_range,
     main = "Toronto Hourly Temperature: May 1–7 Observed + May 8 Forecast",
     xlab = "Date (2003)",
     ylab = "Temperature (°C)",
     xaxt = "n")

# X-axis with day labels
axis(1, at = day_ticks, labels = day_labels, las = 2)

# Forecast (May 8 predicted)
lines(t_future, fc_pred, col = "blue", lwd = 1.5)

# 95% confidence bands
lines(t_future, fc_upper, col = "blue", lty = 2, lwd = 1.2)
lines(t_future, fc_lower, col = "blue", lty = 2, lwd = 1.2)

# Actual May 8 temperature
lines(t_future, may8, col = "red", lty = 2, lwd = 1.5)

legend("topleft",
       legend = c("Observed (May 1–7)", "Forecast (May 8)",
                  "95% CI", "Actual (May 8)"),
       col  = c("black", "blue", "blue", "red"),
       lty  = c(1, 1, 2, 2),
       lwd  = c(1, 1.5, 1.2, 1.5),
       cex  = 0.75, bg = "white")
```

**Commentary:** The ARMA(1,1) model with quadratic trend and 24-hour seasonality captures the daily temperature cycle reasonably well. The confidence bands widen over the 24-hour horizon as uncertainty accumulates. The actual May 8 temperatures (dashed red) fall mostly within the 95% confidence interval, suggesting the model produces a reasonable short-term forecast, though the exact magnitude may drift given that only one week of data was used for estimation.

---

# Problem 3

**Setup.** Let $A_1, A_2$ be uncorrelated, zero-mean RVs with unit variance, and let $\omega \in [0, \pi]$ be fixed. Define
$$
X_t = A_1\cos(\omega t) + A_2\sin(\omega t), \qquad t = 0, \pm 1, \ldots
$$

**Autocovariance.** Since $A_1, A_2$ are uncorrelated with $\operatorname{Var}(A_i)=1$:
$$
\gamma(h) = \operatorname{Cov}(X_t, X_{t+h})
= \cos(\omega t)\cos(\omega(t+h)) + \sin(\omega t)\sin(\omega(t+h))
= \cos(\omega h).
$$

**Find $P_2 X_3$.**

The BLP is $P_2 X_3 = a_1 X_1 + a_2 X_2$ where $\mathbf{a} = \Gamma_2^{-1}\,\mathbf{c}$, with
$$
\Gamma_2 = \begin{pmatrix} \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0) \end{pmatrix}
= \begin{pmatrix} 1 & \cos\omega \\ \cos\omega & 1 \end{pmatrix},
\qquad
\mathbf{c} = \begin{pmatrix} \operatorname{Cov}(X_3, X_1) \\ \operatorname{Cov}(X_3, X_2) \end{pmatrix}
= \begin{pmatrix} \cos(2\omega) \\ \cos(\omega) \end{pmatrix}.
$$

The determinant is $\det(\Gamma_2) = 1 - \cos^2\omega = \sin^2\omega$, so
$$
\Gamma_2^{-1} = \frac{1}{\sin^2\omega}\begin{pmatrix} 1 & -\cos\omega \\ -\cos\omega & 1 \end{pmatrix}.
$$

**Coefficient $a_1$:**
$$
a_1 = \frac{\cos(2\omega) - \cos^2\omega}{\sin^2\omega}.
$$
Using $\cos(2\omega) = 2\cos^2\omega - 1$:
$$
\cos(2\omega) - \cos^2\omega = \cos^2\omega - 1 = -\sin^2\omega
\implies a_1 = -1.
$$

**Coefficient $a_2$:**
$$
a_2 = \frac{-\cos\omega\,\cos(2\omega) + \cos\omega}{\sin^2\omega}
= \frac{\cos\omega\,(1 - \cos(2\omega))}{\sin^2\omega}
= \frac{\cos\omega \cdot 2\sin^2\omega}{\sin^2\omega} = 2\cos\omega.
$$

Therefore:
$$
\boxed{P_2 X_3 = -X_1 + 2\cos(\omega)\, X_2.}
$$

**MSE:**
$$
\operatorname{MSE}(P_2 X_3) = \gamma(0) - \mathbf{c}^\top \mathbf{a}
= 1 - \bigl[\cos(2\omega)\cdot(-1) + \cos(\omega)\cdot 2\cos(\omega)\bigr]
= 1 + \cos(2\omega) - 2\cos^2\omega.
$$
Since $\cos(2\omega) = 2\cos^2\omega - 1$:
$$
\operatorname{MSE}(P_2 X_3) = 1 + (2\cos^2\omega - 1) - 2\cos^2\omega = \boxed{0}.
$$

**Interpretation.** The MSE is zero: knowing $X_1$ and $X_2$ determines $X_3$ exactly. This is because the process $X_t$ has only two free parameters ($A_1$ and $A_2$), so any two observations (at distinct times with $\omega \ne 0, \pi$) completely determine all future values.

---

# Problem 4

**Model.** $X_t = Z_t + \theta Z_{t-1}$, $\{Z_t\} \sim \mathrm{wn}(0,\sigma^2)$. The autocovariances are:
$$
\gamma(0) = (1+\theta^2)\sigma^2, \quad \gamma(1) = \theta\sigma^2, \quad \gamma(h) = 0 \text{ for } |h| \ge 2.
$$

## Part a) Find $P(X_3 \mid X_4, X_5)$

The BLP is $P(X_3 \mid X_4, X_5) = a_1 X_4 + a_2 X_5$ where $\mathbf{a} = \Gamma^{-1}\mathbf{c}$:

$$
\Gamma = \begin{pmatrix} \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0) \end{pmatrix}
= \sigma^2\begin{pmatrix} 1+\theta^2 & \theta \\ \theta & 1+\theta^2 \end{pmatrix},
\qquad
\mathbf{c} = \begin{pmatrix} \operatorname{Cov}(X_3, X_4) \\ \operatorname{Cov}(X_3, X_5) \end{pmatrix}
= \begin{pmatrix} \theta\sigma^2 \\ 0 \end{pmatrix}.
$$

The determinant: $\det(\Gamma) = \sigma^4\bigl[(1+\theta^2)^2 - \theta^2\bigr] = \sigma^4(1 + \theta^2 + \theta^4)$.

$$
\Gamma^{-1} = \frac{1}{\sigma^2(1+\theta^2+\theta^4)}\begin{pmatrix} 1+\theta^2 & -\theta \\ -\theta & 1+\theta^2 \end{pmatrix}.
$$

$$
\mathbf{a} = \Gamma^{-1}\mathbf{c}
= \frac{1}{1+\theta^2+\theta^4}\begin{pmatrix} \theta(1+\theta^2) \\ -\theta^2 \end{pmatrix}.
$$

Therefore:
$$
\boxed{P(X_3 \mid X_4, X_5) = \frac{\theta(1+\theta^2)}{1+\theta^2+\theta^4}\,X_4
\;-\; \frac{\theta^2}{1+\theta^2+\theta^4}\,X_5.}
$$

## Part b) MSE of the Linear Predictor

$$
\operatorname{MSE} = \gamma(0) - \mathbf{c}^\top\mathbf{a}
= (1+\theta^2)\sigma^2 - \theta\sigma^2 \cdot \frac{\theta(1+\theta^2)}{1+\theta^2+\theta^4}.
$$

$$
= (1+\theta^2)\sigma^2\left[1 - \frac{\theta^2}{1+\theta^2+\theta^4}\right]
= (1+\theta^2)\sigma^2 \cdot \frac{1+\theta^4}{1+\theta^2+\theta^4}.
$$

$$
\boxed{\operatorname{MSE}\bigl(P(X_3\mid X_4,X_5)\bigr) = \frac{(1+\theta^2)(1+\theta^4)}{1+\theta^2+\theta^4}\,\sigma^2.}
$$

---

# Problem 5 — Graduate Students Only

**Model.** The ARMA(1,1) process:
$$
Y_t - \tfrac{1}{2}Y_{t-1} = W_t + \tfrac{1}{2}W_{t-1}, \qquad W_t \sim \mathrm{wn}(0,\sigma^2).
$$
So $\phi = \tfrac{1}{2}$, $\theta = \tfrac{1}{2}$.

**Autocovariances (textbook p.78).** For ARMA(1,1) with parameters $\phi, \theta$:

$$
\gamma(0) = \sigma^2\,\frac{1 + 2\phi\theta + \theta^2}{1-\phi^2}, \qquad
\gamma(1) = \sigma^2\,\frac{(\phi+\theta)(1+\phi\theta)}{1-\phi^2}, \qquad
\gamma(h) = \phi^{h-1}\gamma(1),\; h\ge 2.
$$

Substituting $\phi = \theta = \tfrac{1}{2}$:

$$
\gamma(0) = \sigma^2\,\frac{1 + 2\cdot\frac{1}{4} + \frac{1}{4}}{1 - \frac{1}{4}}
= \sigma^2\,\frac{\frac{7}{4}}{\frac{3}{4}} = \frac{7\sigma^2}{3},
$$

$$
\gamma(1) = \sigma^2\,\frac{\bigl(\frac{1}{2}+\frac{1}{2}\bigr)\bigl(1+\frac{1}{4}\bigr)}{\frac{3}{4}}
= \sigma^2\,\frac{1 \cdot \frac{5}{4}}{\frac{3}{4}} = \frac{5\sigma^2}{3}.
$$

## Part a) Find $P(Y_2 \mid Y_1)$

The BLP of $Y_2$ given $Y_1$ is:
$$
P(Y_2 \mid Y_1) = \frac{\gamma(1)}{\gamma(0)}\,Y_1 = \frac{5\sigma^2/3}{7\sigma^2/3}\,Y_1
= \boxed{\frac{5}{7}\,Y_1.}
$$

## Part b) Compute $\operatorname{MSE}\bigl(P(Y_2 \mid Y_1)\bigr)$

$$
\operatorname{MSE} = \gamma(0) - \frac{[\gamma(1)]^2}{\gamma(0)}
= \frac{7\sigma^2}{3} - \frac{(5\sigma^2/3)^2}{7\sigma^2/3}
= \frac{7\sigma^2}{3} - \frac{25\sigma^2}{9} \cdot \frac{3}{7\sigma^2}\cdot\sigma^2.
$$

$$
= \frac{7\sigma^2}{3} - \frac{25\sigma^2}{21}
= \frac{49\sigma^2 - 25\sigma^2}{21} = \boxed{\frac{8\sigma^2}{7}.}
$$
